---
title: "ps05-markdown-plot"
author: "Nev Fenelon"
date: "2023-02-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)

getwd()
```
working directory: "C:/Users/Nev/info201/Problem sets/ps05-rmarkdown-plot"

1 LOAD AND CHECK DATA

1. For solving the problems, and answering the questions, create a new rmarkdown document with an appropriate title.
```{r}
gap <- read_delim("gapminder.csv.bz2")
```

2. (2pt) Load data. How many rows/columns do we have?
```{r}
dim(gap)
```
There are 13,055 rows and 25 columns.

3. Print a small sample of data. Does it look OK?
```{r}
sample_n(gap, 3)
```

2 DESCRIPTIVE STATISTICS

1. How many countries are there in the dataset? Analyze all three: iso3, iso2 and name.
```{r}
length(unique(gap$name))
length(unique(gap$iso2))
length(unique(gap$iso3))
```
There are 250 countries with a name, 249 countries with a iso2 code, and 253 countries with an iso3 code.

2. If you did this correctly, you saw that there are more names than iso-2 codes, and there are even more iso3 -codes. What is going on? Can you find it out?
(a)Find how many names are there for each iso-2 code. Are there any iso-2 codes that correspond to more than one name? What are these countries?
```{r}
gap%>%
  filter(is.na(name))%>%
  group_by(name)%>%
  filter(is.na(iso3))%>%
  distinct(iso2)

gap%>%
  group_by(iso2)%>%
  summarize(num = n_distinct(name)) %>%
  arrange(desc(num))

```
There are not any iso2 codes that corresond to more than one country, however there is a country without an iso2 code, which accounts for the one less iso2 amount (249) than the name amount(250)

(b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have
more than one iso3-code? What are these countries?
```{r}
gap%>%
  filter(is.na(name))%>%
  group_by(name)%>%
  filter(is.na(iso2))%>%
  distinct(iso3)
```
3. (2pt) What is the minimum and maximum year in these data?
```{r}
gap %>%
  filter(!is.na(time))%>%
  summarise(max(time))

gap %>%
  filter(!is.na(time))%>%
  summarise(min(time))


```
3 CO2 emissions

1. (2pt) How many missing co2 emissions are there for each year? Analyze both missing CO2
and co2_PC. Which years have most missing data?
```{r}
gap%>%
  group_by(time)%>%
  summarise(val1 = sum(is.na(co2)), val2 = sum(is.na(co2_PC)))%>%
  arrange(desc(val1))
```
2. (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few more countries of your choice. Explain what do you see.
```{r}
totalCo2 <- gap %>%
  filter(!is.na(co2),(!is.na(name)))%>%
  filter(iso3 == "USA"|iso3 == "CHN"|iso3 == "IND"|iso3 == "FRA"|iso3 == "MEX")

ggplot(data = totalCo2,
  aes(time, co2, color=iso3)) +
  geom_point()
```
Most of the plots have a positive relationship between CO2 and time (as time increases, so does co2 levels). However, Mexico and France have a steady slope, suggesting CO2 levels aren't changing over time.

3. (5pt) Now let’s analyze the CO2 emissions per capita (co2_PC ). Make a similar plot of the same countries. What does this figure suggest?
```{r}
totalCo2_PC <- gap %>%
  filter(!is.na(co2_PC),(!is.na(name)))%>%
  filter(iso3 == "USA"|iso3 == "CHN"|iso3 == "IND"|iso3 == "FRA"|iso3 == "MEX")

ggplot(data = totalCo2_PC,
  aes(time, co2_PC, color=iso3)) +
  geom_point()
```
This is no drastic positive correlation between time and CO2 levels per capita. China's CO2 PC levels are the only ones to rise significantly over time. The other countries either hold a steady CO2 PC level, or slightly decrease over time.

4. (6pt) Compute average CO2 emissions per capita across the continents (assume region is the same as continent). Comment what do you see.
```{r}
gap%>%
  filter(!is.na(co2),!is.na(region))%>%
  filter(time %in% c(1960,2016)) %>%
  group_by(region, time)%>%
  summarize(avgCo2_PC = mean(co2_PC))

```
5. Make a barplot where you show the previous results–average CO2 emissions per capita
across continents in 1960 and 2016.
```{r}
gap %>%
  filter(!is.na(co2),!is.na(region))%>%
  filter(time == 1960 | time == 2016) %>%
  group_by(region, time)%>%
  mutate(avgCo2_PC = mean(co2_PC))%>%
  ggplot(aes(x = region, y = avgCo2_PC, fill = factor(time))) +
  geom_bar(stat='identity', position='dodge')
                     
```
6. Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per
capita) in 2016 for each continent? (Assume region is continent).
```{r}
gap%>%
  filter(!is.na(name),!is.na(co2_PC)) %>%
  filter(time == 2016) %>%
  #group_by(name,region)%>%
  arrange(desc(co2_PC))%>%
    group_by(region)%>%
  filter(rank(co2_PC)<=3|rank(desc(co2_PC)) <=3 )%>%
  select(region,name,co2_PC)


```

4 GDP per capita 

1. (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for 1960. Make the point size dependent on the country size, and color those according to the continent. 
```{r}
plot1 <- gap%>%
  filter(!is.na(GDP_PC),!is.na(lifeExpectancy))%>%
  filter(time == 1960)%>%
  group_by(time, GDP_PC)%>%
  ggplot(aes(x = lifeExpectancy, y = GDP_PC)) +
    geom_point(aes(size = totalPopulation, color = region))

plot1
    
```
2. (4pt) Make a similar plot, but this time use 2019 data only.
```{r}
plot2 <- gap%>%
  filter(!is.na(GDP_PC),!is.na(lifeExpectancy))%>%
  filter(time == 2019)%>%
  group_by(time, GDP_PC)%>%
  ggplot(aes(x = lifeExpectancy, y = GDP_PC)) +
    geom_point(aes(size = totalPopulation, color = region))
plot2
```
3. (6pt) Compare these two plots and comment what do you see. How has world developed
through the last 60 years?

While the curve of the slope for each graph stays relatively the same (the higher the GDP PC, the higher the life expectancy), there are many more countries' data taken so there are many more points on the graph. However, the results from each graph show us the world life expectancy has increased even with a relatively unchanged GDP.

4. (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results fit with what do you see on the figures?
```{r}
gap%>%
  filter(!is.na(lifeExpectancy),!is.na(region))%>%
  filter(time %in% c(1960,2019)) %>%
  group_by(region, time)%>%
  summarize(avgLE = mean(lifeExpectancy))
```
Yes. There is increase of life expectancy over time that is consisted with the data plotted on the 1960 and 2019 graphs.

5. (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results in the order of growth. Explain what do you see.
```{r}

gap%>%
  filter(time == 1960 | time == 2019, !is.na(lifeExpectancy))%>%
  #filter(time %in% c(1960,2016)) %>%
   mutate(lifeExpectancy_prev = lag(lifeExpectancy), 
         lifeExpectancy_growth = lifeExpectancy - lifeExpectancy_prev) %>% 
  group_by(region, time)%>%
  select(region, time, lifeExpectancy_prev, lifeExpectancy_growth)

```
6. (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both
histograms on the same graph, see how well you can do it!
```{r}

gap %>%
  filter(!is.na(GDP_PC),
         !is.na(GDP_PC))%>%
  filter(time == 1960 | time == 2019) %>%
  group_by(time)%>%
  #mutate(avgGDP = mean(GDP_PC))%>%
  ggplot(aes(GDP_PC, fill = factor(time))) +
  geom_histogram(binwidth = 5000, position="dodge")+
  labs(x="Average GDP PC")

```
7. (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When
counting from top.)
```{r}
gap%>%
  filter(!is.na(lifeExpectancy), !is.na(name), time ==1960) %>%
  mutate(rankLE = rank(desc(lifeExpectancy)))%>%
  select(iso3, name, rankLE)%>%
  filter(iso3=="USA")

gap%>%
  filter(!is.na(lifeExpectancy), !is.na(name), time ==2019) %>%
  mutate(rankLE = rank(desc(lifeExpectancy)))%>%
  select(iso3, name, rankLE)%>%
  filter(iso3=="USA")

```
8. (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a bit. But we also have more countries in 2019–what about the relative rank divided by the
corresponding number of countries that have LE data in the corresponding year?
```{r}

rows1960 <- gap %>%
  filter(!is.na(lifeExpectancy), !is.na(name), time == 1960)%>%
  nrow()

gap%>%
  filter(!is.na(lifeExpectancy), !is.na(name), time == 1960) %>%
  mutate(rankLE = rank(desc(lifeExpectancy)),
         relrankLE = rankLE/rows1960)%>%
  select(name, iso3, rankLE, relrankLE)%>%
  filter(iso3 == "USA")
           
rows2019 <- gap %>%
  filter(!is.na(lifeExpectancy), !is.na(name), time == 2019)%>%
  nrow()

gap%>%
  filter(!is.na(lifeExpectancy), !is.na(name), time == 2019) %>%
  mutate(rankLE = rank(desc(lifeExpectancy)),
         relrankLE = rankLE/rows2019)%>%
  select(name, iso3, rankLE, relrankLE)%>%
  filter(iso3 == "USA")

```
